Finite Time Analysis of Linear Stochastic Approximation

### 2017-05-07 intro.tex
initial version
updated intro

\section{Introduction}

Stochastic Approximation (SA) algorithms can be expressed in the general form
\begin{align}\label{eq:sa}
\theta_t=\theta_{t-1}+\alpha_t f(\theta_{t-1},\zeta_t),
\end{align}
where $\theta_t\in \R^d$ are the iterates, and $f(\theta_{t-1},\zeta_t)$ of the function $g(\theta_{t-1})=\EE{f(\theta_{t-1})}_{\zeta_t}$, where the expectation is with respect to some noise process $\zeta_t$. At each time step, the algorithm updates the iterates in increments proportional to a step-size $\alpha_t>0$ and the idea is to eventually converge to $\ts\in \R^d$ which is a desired optimum or a fixed point. Due to their inherent online nature SA algorithms appear widely in machine learning applications \cite{konda,bhatnagar,borkar,gtd,td,bach}, where we have to learn from a noisy data stream.\par
The choice of the step-size $\alpha_t$ needs to trade-off learning rate versus filtering noise, so to speak. In general, one can guess that larger step-size lead to faster learning and at the same time smaller step-sizes are better to handle noise. A popular choice of step-size schedule that is known handle the noise $\alpha_t=O(\frac{1}{t})$. However, since the step size are diminishing, it can hurt the learning rate. An alternative suggestion is to use more aggressive step sizes without minding the noise in the iterates $\theta_t$ in \eqref{eq:sa}, however, introduce additional averaging (known as Polyak-Rupert averaging) to filter the noise. The averaged iterates can be given by $\th_t\eqdef\frac{1}{t}\sum_{s=0}^t \theta_t$, and an analysis of the PR-averaging technique is known from \cite{}.\par

@@

where $b_t\in \R^d$ and $A_t\in \R^{\dcd}$ are data dependent quantities that are preseneted to the learner in a sequential manner. Further, in most instances of LSA $A_t$ turns out to be a rank-$1$ matrix and $A_t\theta_{t-1}$ can be computed in $O(n)$, and are desired for this low per iteration computational cost. Recently, \cite{bach} used \eqref{eq:lsaintro} with constant step-size, i.e., $\alpha_t=\alpha,\,\forall t\geq 0, \, \alpha >0$ and PR-averaging for the linear prediction problem with squared loss and $i.i.d$ sampling. A ``remarkable'' result by \cite{bach} is that there exists a constant step-size such that for any problem where the noisy data is bounded by a known constant, he expected squared prediction error after $t$ updates is at most $\frac{C}{t}$ with a constant $C>0$ that depends \emph{only} on the bound on the data.\todoc{Dimension?}

>>>>

\section{Introduction}
\begin{comment}
Stochastic Approximation (SA) algorithms can be expressed in the general form
\begin{align}\label{eq:sa}
\theta_t=\theta_{t-1}+\alpha_t f(\theta_{t-1},\zeta_t),
\end{align}
where $\theta_t\in \R^d$ are the iterates, and $f(\theta_{t-1},\zeta_t)$ of the function $g(\theta_{t-1})=\EE{f(\theta_{t-1})}_{\zeta_t}$, where the expectation is with respect to some noise process $\zeta_t$. At each time step, the algorithm updates the iterates in increments proportional to a step-size $\alpha_t>0$ and the idea is to eventually converge to $\ts\in \R^d$ which is a desired optimum or a fixed point. Due to their inherent online nature SA algorithms appear widely in machine learning applications \cite{konda,bhatnagar,borkar,gtd,td,bach}, where we have to learn from a noisy data stream.\par
The choice of the step-size $\alpha_t$ needs to trade-off learning rate versus filtering noise, so to speak. In general, one can guess that larger step-size lead to faster learning and at the same time smaller step-sizes are better to handle noise. A popular choice of step-size schedule that is known handle the noise $\alpha_t=O(\frac{1}{t})$. However, since the step size are diminishing, it can hurt the learning rate. An alternative suggestion is to use more aggressive step sizes without minding the noise in the iterates $\theta_t$ in \eqref{eq:sa}, however, introduce additional averaging (known as Polyak-Rupert averaging) to filter the noise. The averaged iterates can be given by $\th_t\eqdef\frac{1}{t}\sum_{s=0}^t \theta_t$, and an analysis of the PR-averaging technique is known from \cite{}.\par
\end{comment}

A linear stochastic approximation (LSA) algorithm of the form given by

@@

where the iterates $\theta_t\in \R^d$ and step-sizes $\alpha_t>0$ are quantities related to the algorithm, while $b_t\in \R^d$ and $A_t\in \R^{\dcd}$ is the data sequence presented to the algorithm.
LSA algorithms are useful in machine learning and reinforcement learning (RL) applications, where the aim is to compute a $\ts\in \R^d$ using the noisy data sequence $(b_t,A_t)$. Specific examples include the stochastic gradient descent algorithm for the problem of linear predition with quadratic loss, and the temporal difference class (TD) of learning algorithms in RL.
An additional feature in these class of pplications is that $A_t$ turns out to be a rank-$1$ matrix and $A_t\theta_{t-1}$ can be obtained in $O(d)$, which is attractive due to the cheap per time step computational requirement.\par
It is of interest to understand the effect of the step-sizes (a quantity that can be controlled by the algorithm designer) on the performance of the LSA. The choice of the step-sizes $\alpha_t$ needs to trade-off learning rate versus filtering noise, so to speak. In general, one can guess that larger step-sizes might lead to faster learning and at the same time smaller step-sizes are better at handling the noise. A technique in literature that goes back to \cite{polyak,ruppert} is to use a constant step-size (CS) $\alpha_t=\alpha>0,\forall t\geq 0$  in conjunction with averaging of iterates , known as Polyak-Rupert (PR) averaging to filter the noise. The averaged iterates can be given by $\th_t\eqdef\frac{1}{t}\sum_{s=0}^t \theta_t$. Under the CS-PR technique $\theta_t$ is the internal vairable and $\thh_t$ is the output of the algorithm. Intuitively, the CS-PR technique can be seen to address the learning vs filtering trade-off, where the constant step-size takes care of the learning  and the averaging of the iterates takes care of the filtering. \par
Recently \cite{bach} considered stochastic gradient descent algorithm (SGD)\footnote{SGD is a special of the LSA algorithm in \eqref{eq:lsaintro}.} with CS-PR for the linear prediction problem with squared loss and $i.i.d$ sampling. A ``remarkable'' result by \cite{bach} is that there exists a constant step-size such that for any problem where the noisy data is bounded by a known constant, the expected squared prediction error after $t$ updates is at most $\frac{C}{t}$ with a constant $C>0$ that depends \emph{only} on the bound on the data.\todoc{Dimension?}
\paragraph{Motivation and Challenges:} In this paper, we wish to study the effect of CS-PR in the case of general LSA algorithms of form \eqref{eq:lsaintro}. Our study is motivated by the problem of linear value function approximation from experience replay in a batch setting in RL using TD class of algorithms. An important difference between the linear prediction problem considered by \cite{bach} and the general LSA setting we consider is that in the former has certain special strucutres and the latter does not. One of the special strucutre exploited by \cite{bach} is the fact that $A_t$ are real symmetric matrices. The general LSA setting does not enjoy such an additional structure presents an interesting challenge.\par
\paragraph{Contributions:} We now list our contributions in this paper
\item We show that in the case of general LSA, there exists instance dependent constant step size choice such that the mean squared error (MSE) of the CS-PR procedure given by $\EE{\norm{\thh_t-\ts}^2}$ is at most $O(\frac{1}{t})$. The MSE further can be broken down to a \emph{bias} term (arising due to the initial condition $\theta_0$) that decays at a rate $O(\frac{1}{t^2})$ and a \emph{variance} term (due to the noise) that decays at at rate $O(\frac{1}{t})$.
\item We observe that when the constant step-size is chosen too big it results in instability, however, a conservative choice of the constant step-size affects only the constants in our error bounds without affecting the asymptotic rate (which might not be true with other popular choices of step-sizes). We aruge heuristically that CS-PR procedure can be deployed in a pricinpled manner by using an outer loop that automatically reduces that constant step-size by detecting detects the blow up of iterates.
\paragraph{Highlights:}
 We only assume that $\EE{A_t}$ is Hurwitz, i.e., all its eigenvalues have positive real parts. This is the condition used in prior works by \cite{polyak-judisky,borkarbook}. Eventhough our proof technique resembles that by \cite{bach}, in addition, we make use of Jordon/Cyclic decomposition in a critical way to handle the general Hurwitz assumption. Our results are applicable to wide variety of TD algorithms such as TD(0), GTD/GDT2, TDC.
Further, the error dynamics has a simple interpretation in terms of behaviour of the sum of terms in a geometric series, and hence offers a simple and unified understanding of the various TD algroithms.
 Further, in most instances of LSA $A_t$ turns out to be a rank-$1$ matrix and $A_t\theta_{t-1}$ can be computed in $O(n)$, and are desired for this low per iteration computational cost. Recently, \cite{bach} used \eqref{eq:lsaintro} with constant step-size, i.e., $\alpha_t=\alpha,\,\forall t\geq 0, \, \alpha >0$ and PR-averaging for the linear prediction problem with squared loss and $i.i.d$ sampling. A ``remarkable'' result by \cite{bach} is that there exists a constant step-size such that for any problem where the noisy data is bounded by a known constant, he expected squared prediction error after $t$ updates is at most $\frac{C}{t}$ with a constant $C>0$ that depends \emph{only} on the bound on the data.\todoc{Dimension?}
